---
title: "Offense Pass Completion as a Predictor of NFL Score Differential"
subtitle: "A comparision betweeen "
author: "Joshua Kwon"
thanks: "Project repository available at: [https://github.com/peteragao/MATH261A-project-template](https://github.com/peteragao/MATH261A-project-template)."
date: today
date-format: long
abstract: "Measuring the importance of an NFL team's offense by"
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(ggpubr)

```

# Introduction

Increase in quarterback pay as % of salary cap [@CBS]. Less room for defense or other positions. Is quarterback really that important?

Since an NFL team wins by scoring more than their opponent at the end of a game, this paper uses score-differential as a proxy for how well a team did during a game. Thus a better 

While intuitive, that a better offense should translate to a better chance at winning, games are won not only by proficient scoring, but by competent defending as well.

This paper serves to address intuition of 

# Data

```{r fig-nfl-scatter}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot with offense completion percentage as the predictor values and score-differential as the outcome values "

nfl <- read.csv("../data/nfl_clean.csv")
ggplot(nfl, aes(x = offense_completion_percentage, y = score_differential)) +
  ggtitle("Score-differential vs offense completion percentage")+
  xlab("offense completion percentage")+
  ylab("score-differential")+
  geom_point() + 
  theme_pubr()
```

The `nfl-team-statistics.csv` dataset provided by SCORE Sports Data Repository [@nfldata] contains statistics about the regular season performance for each NFL team from 1999 to 2022. The data was collected using the `nflreadr` package [@nflreadr]in R.

The purpose of this paper is to observe the relationship between offense completion percentage and score-differential and as such these are the two variables taken from the raw dataset provided by [@nfldata].

```{r clean-summary}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot with offense completion percentage as the predictor values and score-differential as the outcome values "

for_summary <- nfl[c( "offense_completion_percentage", "score_differential")]
summary(for_summary)
```


# Methods

I fit the simple linear regression model $$Y_i = \beta_0 +\beta_1 X_i + \varepsilon_i$$ to understand the relationship between our predictor variable, offense completion percentage, and our outcome variable, score-differential. In this model, $X_i$ represents offense completion percentage of the $i$th observation and $Y_i$ represents the score-differential of the $i$th outcome.$\beta_0$ represents the intercept coefficient, what we expect score-differential to be when the offense has a completion percentage of 0%. $\beta_1$ represents the slope coefficient, what we expect the increase in score-differential will be for every percent increase in completion percentage. In this model, we assume the error term, $\varepsilon_i$ to be random with mean 0 and finite variance $\sigma^2$.

I implemented this analysis using the R programming language [@R_language]

# Results

weakness: Quality of receiver. This regression analysis also ignores the quality of the team's defense. Sc

Organize the quarterbacks by pay to see if paying more for quarterbacks improves score differential

# References
