---
title: "Offense Pass Completion as a Predictor of NFL Score Differential"
subtitle: "A comparision betweeen "
author: "Joshua Kwon"
thanks: "Project repository available at: [https://github.com/peteragao/MATH261A-project-template](https://github.com/peteragao/MATH261A-project-template)."
date: today
date-format: long
abstract: "Measuring the importance of an NFL team's offense by"
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(ggpubr)

nfl <- read.csv("../data/nfl_clean.csv")
```


# Introduction
Increase in quarterback pay as % of salary cap

# Data

# Methods

I fit the simple linear regression model $$Y_i = \beta_0 +\beta_1 X_i + \varepsilon_i$$ to understand the relationship between our predictor variable, offense completion percentage, and our outcome variable, score-differential. In this model, $X_i$ represents offense completion percentage of the $i$th observation and $Y_i$ represents the score-differential of the $i$th outcome.$\beta_0$ represents the intercept coefficient, what we expect score-differential to be when the offense has a completion percentage of 0%. $\beta_1$ represents the slope coefficient, what we expect the increase in score-differential will be for every percent increase in completion percentage. In this model, we assume the error term, $\varepsilon_i$ to be random with mean 0 and finite variance $\sigma^2$.

I implemented this analysis using the R programming language [@R_language]




# Results

```{r fig-scatter}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot with offense completion percentage as the predictor values and score-differential as the outcome values "
ggplot(nfl, aes(x = offense_completion_percentage, y = score_differential)) +
  ggtitle("Score-differential vs offense completion percentage")+
  xlab("offense completion percentage")+
  ylab("score-differential")+
  geom_point() + 
  theme_pubr()
```


weakness: Quality of receiver. This regression analysis also ignores the quality of the team's defense. Sc

# References

https://www.cbssports.com/nfl/news/highest-paid-nfl-qbs-by-salary-cap-percentage-top-2024-bargains-long-term-steals-most-expensive-starters/

https://www.nytimes.com/athletic/6576049/2025/08/28/nfl-qb-contracts-mahomes-allen-prescott-burrow/
